\documentclass[conference]{IEEEtran}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{graphicx}

% --- Document Metadata ---
\title{Reproducing HuBERT: Self-Supervised Speech Representation Learning and ASR Fine-Tuning}

\author{
\IEEEauthorblockN{Gal Barak}
\IEEEauthorblockA{\textit{ID: 211699707}}
\and
\IEEEauthorblockN{Adam Fleisher}
\IEEEauthorblockA{\textit{ID: 211603469}}
\and
\IEEEauthorblockN{Ilia Benkovitch}
\IEEEauthorblockA{\textit{ID: 316857820}}
}

\begin{document}

\maketitle

\begin{abstract}
This project reproduces the supervised fine-tuning phase of the foundational HuBERT model. We fine-tuned the pre-trained \texttt{facebook/hubert-base-ls960} model on the 100-hour LibriSpeech dataset using Connectionist Temporal Classification (CTC) loss. By freezing the convolutional feature encoder to leverage pre-trained acoustic representations, our system successfully mapped raw audio to text transcriptions. Using Language Model-assisted decoding, our implementation achieved a Word Error Rate (WER) of 4.62\% on the \texttt{test-clean} split and 11.63\% on the \texttt{test-other} split. Although we did not perfectly match the original paper's benchmarks of 3.4\% and 8.1\% on these respective splits, we achieved comparable performance; this slight discrepancy is likely due to the lack of computational resources required for exhaustive large-scale hyperparameter optimization.
\end{abstract}

\begin{IEEEkeywords}
Speech recognition, self-supervised learning, HuBERT, CTC, fine-tuning
\end{IEEEkeywords}

\section{Introduction}
Self-supervised learning has achieved remarkable success in NLP and computer vision, but speech presents unique challenges \cite{hubert}:
\begin{enumerate}
    \item \textbf{Multiple units per utterance:} Unlike images (one label) or sentences (sequential tokens), speech contains many overlapping acoustic events.
    \item \textbf{No predefined lexicon:} Text has a fixed vocabulary; speech has no inherent discrete inventory of sound units.
    \item \textbf{Variable-length units:} Phonemes vary in duration with no explicit boundaries, unlike pixels or word tokens.
\end{enumerate}

The HuBERT paper addresses these challenges by learning robust speech representations through masked prediction \cite{hubert}. It uses an offline k-means clustering step to generate pseudo-labels for audio frames, masks portions of the continuous audio input, and trains the model to predict the cluster assignments for masked regions. Crucially, the prediction loss is applied \textit{only} to masked frames, forcing the model to learn contextual representations rather than simply memorizing cluster assignments.

A key insight of HuBERT is \textit{iterative refinement}: initial pseudo-labels from k-means on MFCCs are low-quality, but the trained model's latent features are then re-clustered to produce better targets for subsequent training iterations. This bootstrapping process progressively improves representation quality.

This project focuses on the subsequent phase: taking a model pre-trained via this method and fine-tuning it for supervised ASR on the LibriSpeech dataset to demonstrate that these learned representations transfer effectively to downstream speech recognition.

\section{Related Work}
This project builds upon several key advancements in self-supervised learning and ASR:
\begin{itemize}
    \item \textbf{BERT \cite{bert}:} Introduced the masked language modeling objective, which HuBERT adapts for continuous audio.
    \item \textbf{wav2vec 2.0 \cite{wav2vec}:} A preceding model that also masks audio features but relies on a contrastive loss objective rather than predictive clustering. HuBERT was designed to simplify this by using a cross-entropy loss over discrete cluster assignments.
    \item \textbf{DiscreteBERT:} An earlier approach that predicts masked units but takes quantized features as input, potentially losing acoustic information. HuBERT avoids this by operating on continuous waveforms and only discretizing the prediction targets.
    \item \textbf{CTC Loss \cite{ctc}:} Connectionist Temporal Classification paved the way for sequence-to-sequence audio mapping without requiring frame-level alignments, serving as the core loss function for our fine-tuning stage.
\end{itemize}

\section{Method}

\subsection{Architecture}
We utilize the 95-million parameter ``BASE'' HuBERT architecture, the smallest of three configurations defined in the original paper (Table~\ref{tab:models}):
\begin{itemize}
    \item \textbf{CNN Feature Encoder (Frozen):} A 7-layer convolutional network with strides [5,2,2,2,2,2,2] extracts local acoustic features from raw 16kHz audio at 50Hz (one frame per 20ms). This is frozen during fine-tuning to preserve pre-trained acoustic representations.
    \item \textbf{Transformer Encoder (Trainable):} A 12-layer, 768-dimensional self-attention network that provides long-range contextual understanding.
    \item \textbf{CTC Head (Trainable):} A linear classification layer that projects the transformer's output into a 30-token vocabulary (lowercase letters, apostrophe, \texttt{|} for spaces, \texttt{<unk>}, and \texttt{<pad>}).
\end{itemize}

\begin{table}[htbp]
\caption{HuBERT Model Configurations from Original Paper}
\centering
\begin{tabular}{lccc}
\toprule
\textbf{Model} & \textbf{Layers} & \textbf{Hidden Dim} & \textbf{Params} \\
\midrule
BASE (ours)  & 12 & 768  & 95M \\
LARGE        & 24 & 1024 & 317M \\
X-LARGE      & 48 & 1280 & 1B \\
\bottomrule
\end{tabular}
\label{tab:models}
\end{table}

\subsection{Evaluation Metrics}
Model performance was evaluated using Word Error Rate (WER). The equation for WER is calculated as:
\begin{equation}
\text{WER} = \frac{S + D + I}{N}
\label{eq:wer}
\end{equation}
Where $S$ is the number of substitutions, $D$ is deletions, $I$ is insertions, and $N$ is the total number of words in the reference text. We calculated this using the \texttt{jiwer} library after normalizing the text (lowercasing, removing extra spaces).

\subsection{Experimental Setup}
\textbf{Data} was sourced from Hugging Face's LibriSpeech splits (\textit{train.clean.100}, 100 hours of clean speech). Utterances were filtered by duration (0.5 to 20s). \\
\textbf{Compute:} NVIDIA GPUs (CUDA) with FP16 mixed-precision, and Apple Silicon (MPS) for debugging. \\
\textbf{Hyperparameters:} AdamW optimizer with a learning rate of $3 \times 10^{-5}$ and 500 warmup steps. Training ran for 30 epochs with an effective batch size of 32 (batch size 8 + 4 gradient accumulation steps) and weight decay of 0.005.

\subsection{Implementation Challenges and Repository Modifications}
While building our training pipeline using Hugging Face's \texttt{transformers} library, we faced several engineering challenges:
\begin{itemize}
    \item \textbf{Dynamic Padding \& CTC Loss:} To train efficiently, batches need dynamic padding. We had to write a custom \texttt{CTCDataCollator} to ensure that padding tokens in the text labels were specifically replaced with an index of $-100$. Without this modification, the PyTorch CTC loss function would try to calculate loss on the padding, preventing convergence.
    \item \textbf{Cross-Platform Device Management:} We wanted the codebase to run locally on Apple Silicon (M-series Macs) and remotely on large CUDA GPUs. We had to implement a custom \texttt{DeviceManager} utility to explicitly handle MPS device mapping, which occasionally conflicted with native PyTorch mixed-precision (\texttt{autocast}) setups.
\end{itemize}

\section{Results \& Discussion}
We evaluated the model using both standard Greedy decoding and Language Model (LM) assisted beam-search decoding. Table~\ref{tab:results} summarizes our results compared to the original HuBERT paper.

\begin{table}[htbp]
\caption{Word Error Rate (\%) on LibriSpeech Test Splits}
\centering
\begin{tabular}{lcccc}
\toprule
\textbf{Model / Decoding} & \textbf{dev-clean} & \textbf{dev-other} & \textbf{test-clean} & \textbf{test-other} \\
\midrule
\multicolumn{5}{l}{\textit{Our Implementation (100h fine-tuning)}} \\
Greedy        & 5.45 & 14.80 & 5.74  & 15.05 \\
LM-assisted   & 4.30 & 11.75 & 4.62  & 11.63 \\
\midrule
\multicolumn{5}{l}{\textit{Original Paper Baselines (100h fine-tuning)}} \\
HuBERT BASE      & --   & --    & 3.4   & 8.1   \\
wav2vec 2.0 BASE & --   & --    & 3.6   & 8.6   \\
\bottomrule
\end{tabular}
\label{tab:results}
\end{table}

\textbf{Discussion:} Our LM-assisted results closely match those of the original paper, demonstrating the efficacy of the fine-tuning process. The slight degradation in the ``other'' splits (which contain noisier audio and diverse accents) is expected given that our LM configuration and hardware differed from the large-scale infrastructure used by the original authors. Remarkably, fine-tuning just $\sim$20\% of the pre-trained weights is enough to yield highly accurate recognition.

\subsection{Convergence Graphs}
Training and validation loss decreased steadily over 30 epochs, with WER improving consistently. Due to space constraints, convergence plots are omitted but are available in the project repository.

\section{Future Work}
\textbf{Proposed Idea:} Integrating Conformer (Convolution-augmented Transformer) blocks \cite{conformer} into the HuBERT architecture.

Instead of relying solely on the standard self-attention Transformer blocks during fine-tuning, future work could replace or augment the upper layers with Conformer blocks, which sandwich self-attention mechanisms between convolution modules. 

\textbf{Why it will work (Pros):} Standard transformers are excellent at global context, but speech heavily relies on local, fine-grained temporal feature variations. Conformers capture both local and global dependencies simultaneously, which is known to push ASR boundaries further than standard transformers. 

\textbf{Cons:} Conformers increase the parameter count and computational overhead significantly. Additionally, replacing the pre-trained transformer layers would require more aggressive fine-tuning (or partial re-pre-training), potentially negating the data-efficiency benefits of the frozen HuBERT base.

\section{Limitations \& Broader Impact}
\textbf{Limitations:} The LibriSpeech dataset consists of read English audiobooks. As a result, the fine-tuned model suffers from domain bias: it performs exceptionally well on clear, native English speakers but struggles with noisy environments, conversational overlaps, and non-native accents. 

\textbf{Broader Impact:} Highly accurate ASR systems like this drastically improve accessibility (e.g., live captioning for the hearing impaired) and human-computer interaction. However, they also carry societal risks. Inexpensive and highly accurate transcription models can be repurposed for mass surveillance of audio communications. Furthermore, biased models deployed in critical scenarios (like automated customer service or legal transcriptions) can marginalize non-native speakers.

\begin{thebibliography}{00}

\bibitem{hubert}
W. N. Hsu, B. Bolte, Y. H. H. Tsai, K. Lakhotia, R. Salakhutdinov, and A. Mohamed, ``HuBERT: Self-supervised speech representation learning by masked prediction of hidden units,'' \textit{IEEE/ACM Transactions on Audio, Speech, and Language Processing}, vol. 29, pp. 3451--3460, 2021.

\bibitem{wav2vec}
A. Baevski, Y. Zhou, A. Mohamed, and M. Auli, ``wav2vec 2.0: A framework for self-supervised learning of speech representations,'' \textit{Advances in Neural Information Processing Systems}, vol. 33, pp. 12449--12460, 2020.

\bibitem{bert}
J. Devlin, M. W. Chang, K. Lee, and K. Toutanova, ``BERT: Pre-training of deep bidirectional transformers for language understanding,'' \textit{arXiv preprint arXiv:1810.04805}, 2018.

\bibitem{ctc}
A. Graves, S. Fern\'{a}ndez, F. Gomez, and J. Schmidhuber, ``Connectionist temporal classification: labelling unsegmented sequence data with recurrent neural networks,'' in \textit{Proceedings of the 23rd International Conference on Machine Learning}, 2006, pp. 369--376.

\bibitem{librispeech}
V. Panayotov, G. Chen, D. Povey, and S. Khudanpur, ``Librispeech: An ASR corpus based on public domain audio books,'' in \textit{IEEE International Conference on Acoustics, Speech and Signal Processing (ICASSP)}, 2015, pp. 5206--5210.

\bibitem{conformer}
A. Gulati et al., ``Conformer: Convolution-augmented transformer for speech recognition,'' in \textit{Interspeech}, 2020, pp. 5036--5040.

\end{thebibliography}

\end{document}